Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-02 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: upd figs

@@

We can now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm CS}$, and
gather the terms proportional to $\ell_{\rm CS}$.
%
Doing so, we find,
%
\begin{align}
    \Mf \, \omega^{\rm CS}_{220} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm CS} / \Mf)^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm CS} / \Mf)^4 \right] \, \chi\,,
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$ and,
reassuringly, the nonrotating and linear-in-spin coefficients above agree with those
of~\cite{Maselli:2019mjd} to xx\%.
%
We now isolate the dCS corrections and compare against Eq.~\eqref{eq:qnm_nongr_iso}.
%
Doing so, we can make the identifications,
%
\begin{equation}
p_{\rm CS} = 4\,, \quad \delta \omega^{(0)}_{\rm CS} = 3.1964\,, \quad \delta \omega^{(1)}_{\rm CS} = 41.199\,.
\label{eq:cs_omega_coefs}
\end{equation}
%
The same procedure can be carried for $\tau^{(i)}$, which results in,
%
\begin{equation}
\delta \tau^{(0)}_{\rm CS} = 6.3619\,, \quad \delta \tau^{(1)}_{\rm CS} = 794.66\,.
\label{eq:cs_tau_coefs}
\end{equation}
%
which completes the set of fixed parameters beyond-GR parameters in the ringdown of the \pSEOB{} waveform model.
%
We remark that the alarmingly large values of $\delta \omega_{\rm CS}^{(1)}$ and $\delta \tau_{\rm CS}^{(1)}$
are compensated by the assumptions that $(\ell_{\rm CS} / M_f)^2$ and $\chi$ are much less than unity. These
are the assumptions use to calculate the QNM frequencies both in Ref.~\cite{Wagle:2021tam} and also in
other gravity considered here~\cite{Pierini:2021jxd,Cano:2021myl}.
%
As we said before, the only free beyond-GR parameter left is now $\ell_{\rm CS}$.

@@

\begin{table*}[th]
\begin{tabular}{c | c c c c c c c c}
\hline
\hline
Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
\hline
EdGB      & 4 & 0.0107 & 0.0044 & \dots & \dots &  \cite{Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & \dots \\
dCS      & 4 & 3.1964 & 6.3619 & \dots & \dots &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm CS} \leqslant 34.5$~km \\
cEFTofGR & 4 & $-0.3680$ $(-0.5131)$  & 1.9242 (1.7000) & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
% qEFTofGR & 6 & \dots & \dots & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
\hline
\hline
\end{tabular}
\caption{Summary of the quasinormal modes calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
references from which we used the results from and the current best constraint
(if applicable).
%
In the entries for cEFTofGR, the first entries in the $\delta
\omega^{(i)}_{220}$, $\delta \tau^{(i)}_{220}$ are obtained by considering
only parity-even corrections and the second entries by considering
only parity-odd corrections.
%
In these cases, the dimensionless parameters $\lambda_{\rm e,o}$ are degenerate
with the length scale $\ell$.
%
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

There is a second criteria we must also consider.
%
As already emphasized in Ref.~\cite{Maselli:2019mjd}, the {\sc Parspec}
formalism is by construction perturbative. This means that the non-GR
deformation parameters are small, i.e.,
%
\begin{equation}
\gamma \, \delta \omega^{(j)} \ll 1 \,,
\quad \textrm{and} \quad
\gamma \, \delta \tau^{(j)} \ll 1 \,.
\label{eq:parspec_bound}
\end{equation}
%
for all $j$ of the expansion in dimensionless spin $\chi$.
%
We will also construct posterior distributions on the parameters and check if
most of their support is concentrated to a domain with values much smaller than
unity.

@@

If we find that our posterior obey both conditions~\eqref{eq:eft_bound}
and~\eqref{eq:parspec_bound}, we combine individual posteriors on $\ell$ from
each event.
%
% When possible, we also combine the marginalized posterior distribution on
% $\ell$ for both event GW150914 and GW200129.
%
To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is based on obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. These are then
multiplied together to yield the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/edgb_cdf_varying_threshold.pdf}
\caption{(Color online).~The cumulative distribution function evaluated at the
cut-off $\Lambda_{\rm EFT}(\varepsilon,\gm)$ EdGB gravity as a function of the parameter
$\varepsilon$ for both (blue curves) $\gm = m_2$, the secondary's source
mass, and (orange curves) $\gm = M_{f}$, the remnant's source mass, without
(solid curve) and with (dashed curves) linear in spin QNM corrections. The
horizontal lines lay at 0.68 and 0.90.
}
\label{fig:dcs_cdf}
\end{figure}

@@

%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discussion}
%%%%%%%%%%%%%%%%%

\hs{Try to draw some conclusions that apply to all theories.}

\hs{I think an important message of the paper is that we give indication that
there are theories of gravity (such as dCS) which can by-pass observational
constraints from the inspiral phase alone, {\it yet} they do not, if we
analyze the ringdown. This is quite important because I've frequently heard
that the ``inspiral is what will constrain theories''. I think this might be
the most important message.}

\hs{Say that the fact we spin correction look important reinforces the
importance of calculation QNMs of spinning black holes beyond-GR.}

\hs{Say it is important to go beyond mode calculation and actually understand
how the QNM appear in the GW signal.}

>>>>

% ---------------------------------------------------
% Template. Complete when we have the main text done.
% ---------------------------------------------------
% This paper is organized as follows.
% %
% In Sec.~\ref{sec:review_theories} we present a short overview of the modified gravity theories
% which we will consider and (if any) the current constraints on them.
% %
% In Sec.~\ref{sec:method} we review the two frameworks we combine in our work: the {\sc Parspec}
% parametrization for the QNMs and the {\sc pSEOBNR} waveform model. We also explain how the
% theory-specific QNM calculation can be mapped onto the theory-agnostic {\sc Parspec} framework.
% %
% In Sec.~\ref{sec:results} we explain how we carry ...
% %
% Finally, we present our conclusion in Sec.~\ref{sec:discussion} and point directions for future work.
% %
% Unless stated otherwise, we use geometrical unit $c = G = 1$.

@@

We can now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm CS}$, and
gather the terms proportional to $\ell_{\rm CS}$.
%
Doing so, we find,
%
\begin{align}
    \Mf \, \omega^{\rm CS}_{220} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm CS} / \Mf)^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm CS} / \Mf)^4 \right] \, \chi\,,
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$ and,
reassuringly, the nonrotating and linear-in-spin coefficients above agree with those
of~\cite{Maselli:2019mjd} to xx\%.
%
We now isolate the dCS corrections and compare against Eq.~\eqref{eq:qnm_nongr_iso},
whcih yields,
%
\begin{equation}
p_{\rm CS} = 4\,, \quad \delta \omega^{(0)}_{\rm CS} = 3.1964\,, \quad \delta \omega^{(1)}_{\rm CS} = 41.199\,.
\label{eq:cs_omega_coefs}
\end{equation}
%
The same procedure can be carried for $\tau^{(i)}$, which results in,
%
\begin{equation}
\delta \tau^{(0)}_{\rm CS} = 6.3619\,, \quad \delta \tau^{(1)}_{\rm CS} = 794.66\,.
\label{eq:cs_tau_coefs}
\end{equation}
%
which completes the set of fixed parameters beyond-GR parameters in the ringdown of the \pSEOB{} waveform model.
%
We remark that the alarmingly large values of $\delta \omega_{\rm CS}^{(1)}$ and $\delta \tau_{\rm CS}^{(1)}$
are compensated by the assumptions that $(\ell_{\rm CS} / M_f)^2$ and $\chi$ are much less than unity. These
are the assumptions use to calculate the QNM frequencies both in Ref.~\cite{Wagle:2021tam} and also in
other gravity theories considered here~\cite{Pierini:2021jxd,Cano:2021myl}.
%
As we said before, the only free beyond-GR parameter left is now $\ell_{\rm CS}$.

@@

\begin{table*}[th]
\begin{tabular}{c | c c c c c c c c}
\hline
\hline
Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
\hline
EdGB      & 4 & 0.0107 & 0.0044 & $-$0.2480 & $-$1.101 &  \cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS       & 4 & 3.1964 & 6.3619 & 41.199 & 794.66 &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm CS} \leqslant 34.5$~km \\
cEFTofGR  & 4 & $-0.3680$ $(-0.5131)$  & 1.9242 (1.7000) & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
% qEFTofGR & 6 & \dots & \dots & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
\hline
\hline
\end{tabular}
\caption{Summary of the quasinormal modes calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
references from which we used the results from and the current best constraint
(if applicable).
%
In the entries for cEFTofGR, the first entries in the $\delta
\omega^{(i)}_{220}$, $\delta \tau^{(i)}_{220}$ are obtained by considering
only parity-even corrections and the second entries by considering
only parity-odd corrections.
%
In these cases, the dimensionless parameters $\lambda_{\rm e,o}$ are degenerate
with the length scale $\ell$.
%
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

There is a second criteria we must also consider.
%
As already emphasized in Ref.~\cite{Maselli:2019mjd}, the {\sc Parspec}
formalism is by construction perturbative. This means that the non-GR
deformation parameters are small, i.e.,
%
\begin{equation}
\gamma \, \delta \omega^{(j)} \ll 1 \,,
\quad \textrm{and} \quad
\gamma \, \delta \tau^{(j)} \ll 1 \,.
\label{eq:parspec_bound}
\end{equation}
%
for all $j$ in the expansion in dimensionless spin $\chi$.
%
We will also construct posterior distributions for these parameters and check if
most of their support is concentrated to a domain with values much smaller than
unity.

@@

If we find that our posterior obey both conditions~\eqref{eq:eft_bound}
and~\eqref{eq:parspec_bound}, we combine individual posteriors on $\ell$ from
each event.
%
% When possible, we also combine the marginalized posterior distribution on
% $\ell$ for both event GW150914 and GW200129.
%
To do so, we employ one of the approaches of Ref.~\cite{Perkins:2021mhb} which
is relies on first obtaining a kernel density estimator (KDE) as an approximation for
the posterior distribution on $\ell$ for each individual events. Next, we multiply
these posteriors together to obtain the cumulative posterior.
%
(See also Refs.~\cite{DelPozzo:2011pg,Cardenas-Avendano:2019zxd,Carullo:2021dui}.)
%
This method can have difficulties in dealing with hard cut-offs in the
distribution, such as the one we impose at $\ell = 0$. We address this issue as
done in Ref.~\cite{Perkins:2021mhb}, by first doubling our number of samples as
$\{ s_i \} \cup \{ - s_i \}$, i.e., mirroring the samples across $\ell = 0$.
%
We then renormalize the associated KDE numerically in the range $\ell = [0,
\infty]$.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/edgb_cdf_varying_threshold.pdf}
\caption{(Color online).~The cumulative distribution function evaluated at the
cut-off $\Lambda_{\rm EFT}(\varepsilon,\gm)$ EdGB gravity as a function of the parameter
$\varepsilon$ for both (blue curves) $\gm = m_2$, the secondary's source
mass, and (orange curves) $\gm = M_{f}$, the remnant's source mass, without
(solid curve) and with (dashed curves) linear in spin QNM corrections. The
horizontal lines lay at 0.68 and 0.90. We see that in no situation the curves pass
through the 0.68 or 0.90 lines even at $\varepsilon = 1$. This means that no bound on $\ell_{\rm GB}$
can be placed with the event we analyzed.
}
\label{fig:dcs_cdf}
\end{figure}

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
Event & $j$ & Eq.~\eqref{eq:eft_bound}? & Eq.~\eqref{eq:parspec_bound}?  & Bound ($\gm = M_{f}$) \\
\hline
GW150914 & $j = 0$    & No  & Yes & -- \\
GW150914 & $j = 0, 1$ & No  & Yes & -- \\
GW200129 & $j = 0$    & No  & Yes & -- \\
GW200129 & $j = 0, 1$ & No  & Yes & -- \\
Combined & $j = 0$    & --  & Yes & -- \\
Combined & $j = 0, 1$ & --  & Yes & -- \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for EdGB gravity for GW150914, GW200129, and
combined events using $\gm = M_{f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
%
We conclude that we cannot place any constraint on $\ell_{\rm GB}$ with any of the GW events
studied here.~\hs{TODO: update when completed.}
}
\label{tab:summary_edgb}
\end{table}

@@

%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%

% -------------------------------------------------------------
% \hs{Try to draw some conclusions that apply to all theories.}
% -------------------------------------------------------------

We presented an unified framework to combine the parametrized formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model.
%
We showed through concrete examples how theory-specific calculations QNM of
slowly-rotating BHs in modified gravity theories can be mapped upon the
theory-agnostic beyond-GR parameters of the {\sc ParSpec} formalism.
%
Together this allowed us to apply the \pSEOB~waveform model to test three
modified gravity theories (EdGB, dCS and cubic EFTofGR) using real
observational data from the GW events GW150914 and GW200129.
%
We found, in particular that, within the assumptions of our model and by
stacking posteriors, that the coupling constant of dCS gravity is bound to
$\ell_{\rm CS} \leqslant 34.5$~km.
%
In contrast, we could place any constraint of the coupling constants in EdGB gravity $\ell_{\rm GB}$.
%
This dichotomy between these theories has an analogy with works that considered the inspiral part of the GW signal alone.
%
There it is dCS which is unconstrained, while EdGB, in principle can be constrained, based on results for the closely
related shift-symmetric version of this theory.

Our work shows that there are theories of gravity (such as dCS) which can by-pass observational
constraints from the inspiral phase alone, {\it yet} would not, if we
analyze the ringdown.
%
We hope that our result thus serve as additional motivation for future work
modeling the late-inspiral, merger and ringdown of coalescing BHs in modified
gravity theories.

Let us discuss some avenues for future work \dots

% -----------------------------------------------------------------------------
% \hs{Say that the fact we spin correction look important reinforces the
% importance of calculation QNMs of spinning black holes beyond-GR.}
%
% \hs{Say it is important to go beyond mode calculation and actually understand
% how the QNM appear in the GW signal.}
% -----------------------------------------------------------------------------
